Measuring Hash Trustworthiness via Collision Utility Metrics: Logical Cryptanalysis of MD4

### 2017-12-03 dsn-2018/bare_conf.tex
Extended citations
Signed-off-by: Alexander Scheel <alexander.m.scheel@gmail.com>

From an attacker's perspective, finding a preimage or second preimage has been
difficult: the current published bound for a preimage attack in MD5 is a cost
of $2^{123.4}$ by Y. Sasaki and K. Aoki \cite{SasakiPreimage}. Further, while
J. Kelsey and B. Schneier proposed a method for finding second preimages
in less than $2^n$, we note that this requires rather long messages
\cite{SchneierSecondPreimage}. However, collision attacks are well within reach
of adversaries, and well past theoretical attacks. The work of
H. Dobbertin \cite{Dobbertin1998}, X. Wang \cite{cryptoeprint:2004:199},
M. Schl{\"a}ffer \cite{Schlaffer2006} and others demonstrate the ease with which
collisions can be found.

@@

\textbf{TODO - Finish Organization}
The remainder of this paper is organized as fol
lows. In Section
\ref{Sec:Related} we discuss previous results and show how our research
expands upon prior work. In Section \ref{Sec:Terminology}, we discuss the
terminology and notations we use throughout the remainder of the paper.
In Section \ref{Sec:Intuition}, we give the intuition between measuring
the utility of a collision and motivate why it is useful. In Section
\ref{Sec:Logical}, we present new techniques in logical cryptanalysis to
analyze collisions in hash functions.

@@

We make use of the \textit{bc2cnf} utility by T. Junttila as the basis for
our models \cite{circuits}. This allows us to encode hash functions as generic
structures in the circuit description language and ease the algorithmic
creation of models which build on top of them. We then use CryptoMiniSat5 by
M. Soos to run the models and check for a satisfying witness, often an
example collision \cite{CryptoMiniSat5}. We deviate from evaluating the
performance of SAT solvers or encodings of models by instead focusing on
evaluating the properties of the hash function. We have developed new
techniques for using a SAT solver to inductively build full collisions in MD4
from collisions in reduced-round MD4 and to build addition examples of
collision classes from a single instance of a collision in MD4.

@@

\subsection{Conventions}
\begin{itemize}
    \item We assume ordered tuples and strings are indexable via square
        brackets. Occasionally we use subscripts for referring to
        elements of tuples with named members.
\end{itemize}

@@

Thus, we seek to evaluate the relative difference between a collision
attack and second preimage attacks, to give some measure of distance between
them. Several high profile attacks against hash function usage have been
made, including \textbf{TODO -- cite}, with MD5, cloning a CA certificate
and creating malicious executables, and with SHA-1, a pair of PDFs with the
same hash. However, many of these techniques exploit not the hash, but the
flexibility of the file format, relying instead on embedding arbitrary,
collidable data and later checking for the presence of one of the matching
blocks. This is especially obvious in the MD5 executable and SHA-1 PDF attacks.
In some cases, such as trying to produce a rogue CA certificate that matches
an existing root, this is not possible and thus relies on searching a wide
enough set of objects for a possible match to the collision.

@@

\section{Empirical Results:\\The Utility of Existing Collisions}

@@

\subsection{New Differentials}

We present the following new differentials as part of our work:

\textbf{TODO - figures and graphics on new differentials/differential paths}

\section{ASCII Collisions}

\subsection{Dobbertin}

Default IV

Block 1:

3f474422
624f2b46
683b6c7e
48227d5c
4f2a4f61
206c573e
61622234
227a2633
53482622
706a6727
503e2b3e
7a636c76
7\textbf{b}752620
3f2e5728
236a6376
45654a75

Block 2:

3f474422
624f2b46
683b6c7e
48227d5c
4f2a4f61
206c573e
61622234
227a2633
53482622
706a6727
503e2b3e
7a636c76
7\textbf{c}752620
3f2e5728
236a6376
45654a75

\subsection{Kasselman}

Default IV

Block 1:

7e273957
597e7f4a
7c773d7c
4034376f
69643257
515f5269
6a4d665a
29416b3a
7f377a64
7e232455
6d436c56
26717321
3\textbf{6}65774f
6f2f4421
77787656
26236479

7e273957
597e7f4a
7c773d7c
4034376f
69643257
515f5269
6a4d665a
29416b3a
7f377a64
7e232455
6d436c56
26717321
3\textbf{7}65774f
6f2f4421
77787656
26236479

>>>>

From an attacker's perspective, finding a preimage or second preimage has been
difficult: the current published bound for a preimage attack in MD5 is a cost
of $2^{123.4}$ by Y. Sasaki and K. Aoki \cite{SasakiPreimage}. In MD4, the
best second preimage attack by H. Yu et al. gives a bound of $2^{-56}$
likelihood, but uses destructive message modification techniques to achieve
collisions for all other messages \cite{cryptoeprint:2007:206}. Further, while
J. Kelsey and B. Schneier proposed a method for finding second preimages
in less than $2^n$, we note that this requires rather long messages
\cite{SchneierSecondPreimage}. However, collision attacks are well within reach
of adversaries, and well past theoretical attacks. The work of
H. Dobbertin \cite{Dobbertin1998}, X. Wang \cite{cryptoeprint:2004:199},
M. Schl{\"a}ffer \cite{Schlaffer2006} and others demonstrate the ease with which
collisions can be found.

@@

The remainder of this paper is organized as follows. In Section
\ref{Sec:Related} we discuss previous results and show how our research
expands upon prior work. In Section \ref{Sec:Terminology}, we discuss the
terminology and notations we use throughout the remainder of the paper.
In Section \ref{Sec:Intuition}, we give the intuition between measuring
the utility of a collision and motivate why it is useful. In Section
\ref{Sec:Logical}, we present new techniques in logical cryptanalysis to
analyze collisions in hash functions. In Section \ref{Sec:Empirical}, we
discuss the utility of existing collisions under the metrics described in
\ref{Sec:Intuition}. In Section \ref{Sec:Collision}, we discuss new
differential paths and the effectiveness of various techniques presented
in Section \ref{Sec:Logical}. Lastly, in Section \ref{Sec:Impacts}, we
discuss broader impacts of this research.

@@

We make use of the \textit{bc2cnf} utility by T. Junttila as the basis for
our models \cite{circuits}. This allows us to encode hash functions as generic
structures in the circuit description language and ease the algorithmic
creation of models which build on top of them. We then use CryptoMiniSat5 by
M. Soos to run the models and check for a satisfying witness, often an
example collision \cite{Soos:2009:ESS:1575471.1575502, CryptoMiniSat5}. We
deviate from evaluating the performance of SAT solvers or encodings of models
by instead focusing on evaluating the properties of the hash function. We have
developed new techniques for using a SAT solver to inductively build full
collisions in MD4 from collisions in reduced-round MD4 and to build addition
examples of collision classes from a single instance of a collision in MD4.

@@

\subsection{Conventions}
\begin{itemize}
    \item We assume ordered tuples and strings are indexable via square
        brackets. Occasionally we use subscripts for referring to
        elements of tuples with named members.
    \item We use MD4 as specified in RFC 1320 \cite{rfc1320}.
\end{itemize}

@@

Thus, we seek to evaluate the relative difference between a collision
attack and second preimage attacks, to give some measure of distance between
them. Several high profile attacks against hash function usage have been
made. For MD5, these include cloning a CA certificate \cite{Stevens2009}
and creating malicious executables \cite{MD5Executables}. For SHA-1, a pair of
PDFs with the same hash \cite{cryptoeprint:2017:190}. However, many of these
techniques exploit not the hash, but the flexibility of the file format,
relying instead on embedding arbitrary, collidable data and later checking for
the presence of one of the matching blocks. This is especially obvious in the
MD5 executable and SHA-1 PDF attacks. The work by M. Stevens to perform the
CA Collision, on the other hand, was a chosen prefix attack extending
previous work from 2007 \cite{Stevens2007}.

@@

\section{Empirical Results:\\The Utility of Existing Collisions}
\label{Sec:Empirical}

@@
